Universals of word order result from optimization of grammars for efficient communication

### 2019-02-07 second-draft/paper-names.tex
continuing revision

We model the process where a speaker communicates an utterance to a listener (Figure~\ref{fig:comm}).
TODO Shannon
We model the meaning that the speaker communicates as the basic predicate-argument structure.
We represent these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.

By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined; the dependency tree (including the labels on the arcs) specifies how the meanings of words are combined.
Therefore, a listener needs to recover the information provided in the dependency structure in order to understand a sentence correctly.

We can view this as a formalization of the Language of Thought (CITE) TODO explain why

When uttering, the speaker needs to choose an ordering in which to order the words in this tree to generate a string of words.
The listener receives a string of words, and attempts to reconstruct the underlying syntactic relations.
%We formalize the listener's understanding as probabilistic inference, that is, the listener 
If communication is successful, the listener can identify the intended structure with high confidence~(Figure~\ref{fig:comm}).

Speakers encode dependency trees into sentences according to a grammar~(Figure~\ref{fig:grammars}).
While natural languages differ in the freedom of word order, all languages rely on some amount of word order reglarity to encode meaning.
This is illustrated in Figure~\ref{fig:grammars}: Grammars specify how dependency structures are encoded into strings of words.
For instance, Grammar 1 -- corresponding to Arabic in Figure~\ref{fig:arabic-japanese-simple} -- orders objects (`friends', `letter') after verbs and has prepositions (`to friend').
Grammar 2 orders objects after verbs but has postpositions (`friend -- to').
Grammars 3 and 4 place the object before the verb, and one of them (Grammar 3) corresponds to Japanese order.
In a speech community, speakers and listeners have shared knowledge of these rules, which allows the listener to decode the dependency tree.
In the terminology of information theory, these rules specify a \emph{channel} through which the speaker transmits the dependency structures.

Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance (TODO can we find an example for this?)

@@

All natural languages have some degree of word order regularities that are specified in their grammar.
It is thus not realistic to assume that speakers optimize each utterance individually.
Instead, we hypothesize that the word order rules of languages as a whole are optimized for efficiency, and that this optimization produces the observed universals.

We formalize a language as a distribution over dependency structures $T$ together with an ordering grammar $\theta$ that specifies how dependency structures are ordered into strings of words.
Applying the ordering grammar to the dependency structures, a language defines a probability distribution over sentences $L$.
Efficiency is defined as a tradeoff between

@@

We create

Fortunately, we have exacly the kind of data we need

For the dependency structures, we take real tree structures from corpora of 51 languages.
These corpora represent a typologically and genetically diverse group of languages.

For each of these corpora, we construct grammars that optimize efficiency of the linearized sentences.
(The original word orders of the languages do not enter.)

See SI for our method for creating optimized grammars.

@@

\section{DLM}

>>>>

We model the process where a speaker communicates an utterance to a listener (Figure~\ref{fig:comm}) in Shannon's framework of information theory.
In Shannon's model, a transmitter encodes a message into a signal. The receiver decodes the signal, attempting to reconstruct the original message.
Applying this model to word order, we take the message to consist of a set of syntactic and semantic relations.
Following a long tradition in formal and computational liguistics, we formalize these in the format of Dependency Grammar:
This linguistic formalism draws directed arcs between syntactically related words, annotated with syntactic relations.
In Figure~\ref{fig:comm}, the message consists of a dependency structure with the words `saw', `Mary', `John', where Mary is the subject of the event denoted by `saw', and John is the object.

When uttering, the speaker needs to choose an ordering in which to order the words in this tree to generate a string of words.
The listener receives this string of words.
By the principle of compositionality \cite{frege1892sinn}, the meaning of a sentence is a function of the meanings of the parts and how they are combined.
The dependency structure (including the labels on the arcs) specifies how the meanings of words are combined.
Therefore, a listener needs to recover the information provided in the dependency structure in order to understand a sentence correctly.
Consistent with Shannon's model, we assume a Bayesian listener who decodes dependency structures probabilistically~(Figure~\ref{fig:comm}).
If communication is successful, the listener can identify the intended structure with high confidence~(Figure~\ref{fig:comm}).

%We can view this as a formalization of the Language of Thought (CITE) TODO explain why

%We formalize the listener's understanding as probabilistic inference, that is, the listener

All natural languages have some degree of word order regularities that are specified in their grammar.
For instance, English places subjects before and objects after the verb, allowing the listener to unambiguously decode the sentence in Figure~\ref{fig:comm}.
That is, speakers and listeners have shared knowledge of a \key{grammar} that specifies how dependency structures are encoded into sentences.
Natural languages differ in the rules they apply: Some place the object after the verb, some place it before the verb.
%Speakers encode dependency trees into sentences according to a grammar~(Figure~\ref{fig:grammars}).
%While natural languages differ in the freedom of word order, all languages rely on some amount of word order reglarity to encode meaning.
This is illustrated in Figure~\ref{fig:grammars}: Grammars specify how dependency structures are encoded into strings of words.
For instance, Grammar 1 -- corresponding to Arabic in Figure~\ref{fig:arabic-japanese-simple} -- orders objects (`friends', `letter') after verbs and has prepositions (`to friend').
Grammar 2 orders objects after verbs but has postpositions (`friend -- to').
Grammars 3 and 4 place the object before the verb, and one of them (Grammar 3) corresponds to Japanese order.
%In a speech community, speakers and listeners have shared knowledge of these rules, which allows the listener to decode the dependency tree.
%In the terminology of information theory, these rules specify a \emph{channel} through which the speaker transmits the dependency structures.

In Information Theory, the usefulness of a communication channel depends on (1) how costly encoding and transmission is, and (2) how precisely messages can be recovered from codes.
In our model, these will depend on the grammar:
Grammars can differ in the degree to which a listener can unambiguously recover the dependency structure from an utterance (TODO can we find an example for this?)

@@

%We formalize a language as a distribution over dependency structures $T$ together with an ordering grammar $\theta$ that specifies how dependency structures are ordered into strings of words.
%Applying the ordering grammar to the dependency structures, a language defines a probability distribution over sentences $L$.
%Efficiency is defined as a tradeoff between

@@

\subsection{Optimizing Grammars for Efficiency}

We now ask:
(1) Do the grammars of natural languages evolve towards optimizing efficiency of communication?
(2) Does this process of optimization account for Greenberg's order correlation universals?

%For answering these questions,

Answering these questions requires a sample of dependency structures as actually used by speakers of difefrent languages.
Such samples have recently become available with the Universal Dependencies project, which has collected and created dependency annotations for several dozens of different languages.
We use data from 51 languages.
These corpora represent a typologically and genetically diverse group of languages.

To answer whether languages evolve to have efficient word order, we compare the efficiency of the actual grammars of these 51 languages to randomly constructed baseline grammars.
To show that this process of efficiency optimization accounts for Greenberg's correlation universals, we computationally construct grammars that optimize efficiency.
We then show that these grammars mostly exhibit the Greenbergian order correlation universals.
We furthermore show that this process of optimization also \emph{explains} DLM, providing a first-principles explanation of this heuristic generalization.

For every one of the 51 corpora, we computationally construct eight optimal grammars.
Note that the original orders of the actual languages do not enter this objective.
See SI for our method for creating optimized grammars.

%
%All natural languages have some degree of word order regularities that are specified in their grammar.
%It is thus not realistic to assume that speakers optimize each utterance individually.
%Instead, we hypothesize that the word order rules of languages as a whole are optimized for efficiency, and that this optimization produces the observed universals.
%

@@

\section{DLM TODO put these at the appropriate place}
